### 2014-08-11 doc/thesis/typedlua.tex
Rewriting the section about atomic types and functions

In the following example we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function fact (n:number):number
      if n == 0 then
        return 1
      else
        return n * fact(n - 1)
      end
    end
    local x = 5
    print(fact(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type $\Number$ to \texttt{x}.
The compiler does not use the literal type $5$ instead of the base
type $\Number$ because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

@@

The function \texttt{dist} receives two parameters of type $\Any$
and returns a value of type $\Number$.
The compiler assigns the dynamic type $\Any$ to the input
parameters of \texttt{dist} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
The compiler infers the return type of \texttt{dist} because it is
not a recursive function, that is, Typed Lua does not infer the return
type of recursive functions.

@@

In the previous example, we cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.

>>>>

In the following example we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function factorial (n:number):number
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type $\Number$ to \texttt{x}, and this example compiles
without any warnings.
The compiler does not use the literal type $5$ instead of the base
type $\Number$ because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

@@

The function \texttt{dist} receives two parameters of type $\Any$
and returns a value of type $\Number$.
The compiler assigns the dynamic type $\Any$ to the input
parameters of \texttt{dist} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotated the return type of \texttt{dist},
the compiler is able to infer its return type because it is local
and not recursive.

@@

In this example, Typed Lua cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

Lua has first-class functions, and they have three peculiarities:
flexible arity, multiple return values, and multiple assignment.
In the first case, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
In the second case, Lua function can return any number of values,
and this number of returned values may not be statically known.
In the third case, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment
(or vice-versa), that is, calling a function is like doing a
multiple assignment where the left hand side is the parameter list
and the right hand side is the argument list.

Typed Lua uses \emph{second-level types} to encode functions with
flexible arity, multiple returns, and multiple assignment.
We call them second-level because these types do not correspond
to actual Lua values and we cannot use them to type variables or
parameters.
They are just an artifact of the interaction between the type system
and the semantics of Lua.

As we can see in Figure \ref{fig:typelang}, a second-level type is
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
A variadic type $T{*}$ is a generator for a sequence of values of
type $T \cup \Nil$.
Unions of tuples play an important role in functions that are
overloaded on the return type, together with \emph{projection types}.
We will explain union types and projection types in more detail
in the next section.

To match the semantics of Lua function calls, the default operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{abs} is
$\Number \times \Value{*} \rightarrow \Number \times \Nil{*}$,
and the type of \texttt{dist} is
$\Any \times \Any \times \Value{*} \rightarrow \Number \times \Nil{*}$.
If we call \texttt{abs} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{abs} receive any number of
extra arguments.
If we call \texttt{abs} with fewer arguments, Typed Lua uses $\Nil$ to
replace the missing arguments and in the case of \texttt{abs} generates
a warning, as $\Nil$ is neither a subtype of $\Number$ nor consistent
with $\Number$.

Typed Lua also provides a strict operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{abs} is $\Number \rightarrow \Number$, and the type of
\texttt{dist} is $\Any \times \Any \rightarrow \Number$.
If we call \texttt{abs} with either extra or missing arguments,
Typed Lua generates warnings about arity mismatch.

A variadic type can only appear in the tail position of a tuple,
because Lua takes only the first value of any expression that appears
in an expression list that is not in tail position.
The following example shows the interaction between multiple returns
and expression lists:
\begin{verbatim}
local function multiple ()
  return 2, "foo"
end
local function sum (x:number, y:number)
  return x + y
end
local x, y, z = multiple(), multiple()
print(sum(multiple(), multiple()))
\end{verbatim}

In the default mode, the type of \texttt{multiple} is
$\Value{*} \rightarrow \Number \times \String \times \Nil{*}$,
and the type of \texttt{sum} is
$\Number \times \Number \times \Value{*} \rightarrow \Number \times \Nil{*}$.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
$\Number \times \Number \times \String \times \Nil{*}$,
and the type assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are respectively $\Number$, $\Number$, and $\String$.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

In the strict mode, the type of \texttt{multiple} is
$\Void \rightarrow \Number \times \String$,
and the type of \texttt{sum} is $\Number \times \Number \rightarrow \Number$.
Like in the default mode, only the first value produced by the
first call to multiple gets used in the right hand side of the
multiple assignment;
and the type of the right hand side is also
$\Number \times \Number \times \String \times \Nil{*}$,
because the strict mode affects only the type of function declarations
and multiple assignments that appear in function calls.
Unlike in the default mode, the type of the call to \texttt{sum} is
$\Number \times \Number \times \String$, and does not include the
variadic tail to catch arity mismatch.
Thus, this example compiles with error because the call to \texttt{sum}
passes an unexpected extra argument of type $\String$.
